\section{Virtual Memory and Paging}
\subsection*{Q1. GATE 2008: Multi-level Paging Table Address Bits}

A processor uses \textbf{36-bit physical addresses} and \textbf{32-bit virtual addresses}, with a page frame size of \textbf{4 KB}. Each page table entry is of size \textbf{4 bytes}. A \textbf{three-level page table} is used for virtual to physical address translation, where the virtual address is used as follows:

\begin{itemize}
    \item Bits \textbf{30--31} index the \textbf{first level} page table.
    \item Bits \textbf{21--29} index the \textbf{second level} page table.
    \item Bits \textbf{12--20} index the \textbf{third level} page table.
    \item Bits \textbf{0--11} are the \textbf{page offset}.
\end{itemize}

The number of bits required for addressing the next level page table (or page frame) in the page table entries of the \textbf{first, second, and third level} page tables are respectively:

\begin{enumerate}[label=(\alph*)]
    \item 20, 20, 20 \hspace{5cm} (c) 24, 24, 20
    \item 24, 24, 24 \hspace{5cm} (d) 25, 25, 24
\end{enumerate}

% -------------------------------------------
\subsection*{Q2. Multi-level Virtual Address Structure}

A system uses \textbf{40-bit virtual addresses} and\\
\textbf{36-bit physical addresses}. The page size is \textbf{8 KB},and each page table\\ 
entry is \textbf{4 bytes}. The virtual address is broken down as follows for a\\
\textbf{2-level page table}:

\begin{itemize}
    \item Bits \textbf{29--39} for the \textbf{first level} index
    \item Bits \textbf{16--28} for the \textbf{second level} index
    \item Bits \textbf{0--15} for the \textbf{page offset}
\end{itemize}

What is the size (in bits) of the physical page number stored in each page table entry?

\begin{enumerate}[label=(\alph*)]
    \item 36 \hspace{5cm} (c) 24
    \item 23 \hspace{5cm} (d) 28
\end{enumerate}

% -------------------------------------------
\subsection*{Q3. Number of Entries per Table}

A processor has a \textbf{48-bit virtual address}, and page size is \textbf{4 KB}. A \textbf{three-level page table} is used with the following split:

\begin{itemize}
    \item Level 1: 12 bits
    \item Level 2: 12 bits
    \item Level 3: 12 bits
    \item Offset: 12 bits
\end{itemize}

How many entries are there in each level of the page table?

\begin{enumerate}[label=(\alph*)]
    \item $2^{12}, 2^{12}, 2^{12}$ \hspace{5cm} (c) $2^{10}, 2^{12}, 2^{14}$
    \item $2^{16}, 2^{16}, 2^{4}$  \hspace{5cm} (d) $2^{16}, 2^{12}, 2^{8}$
\end{enumerate}

% -------------------------------------------
\subsection*{Q4. Page Table Size Estimation}

A 32-bit machine uses a \textbf{2-level paging} scheme with \textbf{4 KB pages}. Virtual address is split as:

\begin{itemize}
    \item Level 1 index: 10 bits
    \item Level 2 index: 10 bits
    \item Offset: 12 bits
\end{itemize}

Assume:
\begin{itemize}
    \item Each page table entry is \textbf{4 bytes}
    \item Only \textbf{1 process is running}
\end{itemize}

What is the \textbf{maximum size} (in bytes) of the total page tables needed for a process?

\begin{enumerate}[label=(\alph*)]
    \item 8 KB  \hspace{5cm} (c) 4 MB
    \item 16 KB \hspace{5cm} (d) 4 KB
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Synchronization %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\input{synchronization.tex}