\section{Verbal Reasoning – Analogy, Grammar, Coherence, Inference}

\subsection*{Q1. Analogy (Vocabulary)}
\textbf{walk : jog : sprint :: bothered : \rule{3cm}{0.1mm} : daunted}

\begin{enumerate}[label=(\alph*)]
    \item phased
    \item phrased
    \item fazed
    \item fused
\end{enumerate}

\textbf{Correct Answer:} (c) fazed  
\textit{Explanation:} Walk, jog, sprint represent increasing intensity. Similarly, bothered → fazed → daunted shows increasing psychological intensity.



\subsection*{Q2. Fill in the Blank (Preposition Usage)}
She looked \rule{2cm}{0.1mm} the list of names to find hers, but it wasn’t there.

\begin{enumerate}[label=(\alph*)]
    \item on
    \item down
    \item for
    \item on
\end{enumerate}

\textbf{Correct Answer:} (b) down  
\textit{Explanation:} "Looked down the list" is the correct phrasal usage.

---

\subsection*{Q3. Sentence Arrangement (Coherent Sequence)}
Arrange the following four sentences in a logical order:

\begin{enumerate}[label=\Roman*.]
    \item She picked up the phone and dialed a number.
    \item A few moments later, she smiled and hung up.
    \item The phone rang three times before someone answered.
    \item Her eyes were anxious as she searched for the contact.
\end{enumerate}

\textbf{Which of the following sequences forms a coherent narrative?}

\begin{enumerate}[label=(\alph*)]
    \item IV, I, III, II
    \item I, III, IV, II
    \item IV, III, I, II
    \item I, IV, II, III
\end{enumerate}

\textbf{Correct Answer:} (a) IV, I, III, II  
\textit{Explanation:} Starts with her searching, then dialing, then ringing, then ending with a smile.

---

\subsection*{Q4. Paragraph Inference}
\textbf{Paragraph:}  
Climate change models predict that rising sea levels and increased temperature variations will lead to changes in the distribution of many marine species. Fish that are normally found in tropical waters may migrate to temperate zones in search of optimal conditions. This could have major consequences for global fisheries and local ecosystems.

\textbf{Which of the following can be inferred with certainty?}

\begin{enumerate}[label=(\alph*)]
    \item Marine biodiversity will increase in temperate zones.
    \item Fish migration patterns are affected by water temperature.
    \item All tropical fish will migrate to cooler waters.
    \item Global fisheries will collapse due to climate change.
\end{enumerate}

\textbf{Correct Answer:} (b)  
\textit{Explanation:} Only option (b) is directly supported and certain based on the paragraph.

