\section{Properties of Determinants: Conceptual Questions}

\subsection*{Q1.}
If the determinant of matrix $A$ is zero, then which of the following must be true?
\begin{enumerate}[label=(\alph*)]
    \item $A$ is invertible
    \item $A$ has linearly independent rows
    \item $A$ is singular
    \item $A^2$ has a non-zero determinant
\end{enumerate}

\subsection*{Q2.}
If $\det(A) = 5$, what is the value of $\det(3A)$ for a $3 \times 3$ matrix?

\subsection*{Q3.}
If $A$ and $B$ are square matrices of the same order, then $\det(AB) = \rule{2cm}{0.15mm}$

\subsection*{Q4.}
If $A$ is a $n \times n$ matrix and $\lambda$ is a scalar, then $\det(\lambda A) = $ ?

\subsection*{Q5.}
True or False: If two rows of a matrix are interchanged, the determinant remains the same.

\subsection*{Q6.}
Which of the following matrices has determinant zero?
\begin{enumerate}[label=(\alph*)]
    \item A matrix with two identical rows
    \item A diagonal matrix with one entry zero
    \item A matrix whose rank is less than its order
    \item All of the above
\end{enumerate}

\subsection*{Q7.}
If $\det(A) = d \neq 0$, then what is $\det(A^{-1})$?

\subsection*{Q8.}
If $A$ is orthogonal, then $\det(A) = $ ?

\subsection*{Q9.}
Let $A$ be a $3 \times 3$ matrix with $\det(A) = 2$. What is $\det(\text{adj}(A))$?

\subsection*{Q10.}
Which of the following operations does **not** change the determinant value?
\begin{enumerate}[label=(\alph*)]
    \item Swapping two rows
    \item Multiplying a row by a scalar
    \item Adding a multiple of one row to another
    \item Multiplying entire matrix by scalar
\end{enumerate}

\subsection*{Q11.}
Let $A$ be a skew-symmetric matrix of odd order. Then $\det(A)$ is:

\subsection*{Q12.}
If $A$ is a $3 \times 3$ matrix with real entries such that $\det(A) = 0$, then which of the following is not necessarily true?
\begin{enumerate}[label=(\alph*)]
    \item $A$ is not invertible
    \item $A$ has a zero eigenvalue
    \item $A$ has full rank
    \item $A$ is singular
\end{enumerate}

\subsection*{Q13.}
True or False: $\det(A) = \det(A^T)$ for any square matrix $A$

\subsection*{Q14.}
Let $A$ be $2 \times 2$ with $\det(A) = 7$. What is the determinant of $A^T A$?

\subsection*{Q15.}
Let $A$ be a diagonal matrix with entries $(\lambda_1, \lambda_2, \dots, \lambda_n)$. What is $\det(A)$?

\subsection*{Q16.}
If $A$ is a square matrix and $\det(A) \neq 0$, then what can we say about the solution of $Ax = b$?

\subsection*{Q17.}
Let $A$ be an upper triangular matrix with non-zero diagonal entries. Then, which of the following is true?
\begin{enumerate}[label=(\alph*)]
    \item $\det(A) = 0$
    \item $\det(A)$ is sum of diagonal entries
    \item $\det(A)$ is product of diagonal entries
    \item $\det(A)$ is not defined
\end{enumerate}

\subsection*{Q18.}
If $\det(A) = 3$ and $\det(B) = 4$, then $\det(AB^{-1}) = $ ?

\subsection*{Q19.}
Let $A$ be a matrix such that $\det(A) = 0$. Which of the following is not guaranteed?
\begin{enumerate}[label=(\alph*)]
    \item $A$ is non-invertible
    \item $A$ has at least one zero eigenvalue
    \item $\text{Rank}(A) < \text{order of } A$
    \item $A$ is diagonalizable
\end{enumerate}

\subsection*{Q20.}
If a matrix $A$ has all its elements in a row or column zero,\\ then $\det(A) = $ ?

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\section[Determinants: Using Properties for Evaluation]
{Determinants:\\ Using Properties for Evaluation}

\subsection*{Q1.}
Let $A$ be a $3 \times 3$ matrix such that $\det(A) = 4$. What is the value of $\det(2A^{-1})$?

\subsection*{Q2.}
Let $A$ be a $4 \times 4$ matrix with $\det(A) = 5$. If $B$ is obtained from $A$ by swapping two rows and multiplying one row by $3$, then what is $\det(B)$?

\subsection*{Q3.}
Let $A$ be a $3 \times 3$ matrix such that $\det(A) = -2$. Let $C = A^2 \cdot (3I)$. Compute $\det(C)$.

\subsection*{Q4.}
If $A$ is a $5 \times 5$ matrix and $A'$ is obtained from $A$ by adding twice the third row to the fifth row, what can be said about $\det(A')$?

\subsection*{Q5.}
Let $A$ be a $3 \times 3$ matrix such that $\det(A) = 6$. Compute $\det(A^T A^{-1})$.

\subsection*{Q6.}
Let $A$ be a $3 \times 3$ skew-symmetric matrix. Compute $\det(A^T A)$.

\subsection*{Q7.}
Let $A$ be an upper triangular $4 \times 4$ matrix with diagonal entries $1, 2, 3, 4$. Compute $\det(2A)$.

\subsection*{Q8.}
Let $A$ be a $4 \times 4$ matrix with $\det(A) = 2$. Matrix $B$ is obtained by multiplying the second and third rows of $A$ by $-1$. Find $\det(B)$.

\subsection*{Q9.}
Let $A$ be a $3 \times 3$ invertible matrix with $\det(A) = -5$. Let $B = A + 3I$. Can $\det(B)$ be computed directly from this information? Justify.

\subsection*{Q10.}
Let $A$ be a $3 \times 3$ matrix with all row sums equal to $5$. What is the value of $\det(A)$ if we subtract the first row from the other two rows?

