\begin{center}
    \Large\textbf{UPSC Mains – GS Paper 1}\\
    \normalsize Chapter 7: \textit{Towards Freedom and Partition (1939–1947)}\\
    \vspace{0.5em}
    \textit{Based on \textbf{From Plassey to Partition} by Sekhar Bandyopadhyay}
\end{center}

\vspace{1em}

\subsection*{Q.1 [Impact of World War II]}
How did the outbreak of World War II alter the course of the Indian freedom struggle?

\subsection*{Q.2 [August Offer and Cripps Mission]}
Evaluate the British efforts like the August Offer and the Cripps Mission to secure Indian cooperation during World War II.

\subsection*{Q.3 [Quit India Movement]}
Critically examine the causes, spread, and impact of the Quit India Movement of 1942.

\subsection*{Q.4 [INA and Subhas Chandra Bose]}
Assess the role of Subhas Chandra Bose and the Indian National Army (INA) in India's struggle for independence.

\subsection*{Q.5 [Post-War Political Developments]}
Trace the political developments between 1945 and 1947 that led to Indian independence.

\subsection*{Q.6 [Labour and Peasant Movements]}
Discuss the rise of labour and peasant movements in the 1940s and their role in the nationalist movement.

\subsection*{Q.7 [Communal Politics and Violence]}
Analyze the rise of communalism in the 1940s and its contribution to the Partition of India.

\subsection*{Q.8 [Cabinet Mission Plan]}
What were the objectives and proposals of the Cabinet Mission Plan (1946)? Why did it ultimately fail?

\subsection*{Q.9 [Direct Action Day and Its Consequences]}
Discuss the significance and consequences of the Direct Action Day (1946).

\subsection*{Q.10 [Mountbatten Plan and Indian Independence Act]}
Examine the key features of the Mountbatten Plan and how it shaped the Indian Independence Act of 1947.

\subsection*{Q.11 [Partition: Causes and Consequences]}
What were the key causes of Partition? Comment on its social, economic, and political consequences.

\subsection*{Q.12 [Role of Congress in Negotiations]}
How did the Indian National Congress respond to British proposals during the 1940s?

\subsection*{Q.13 [Muslim League’s Demands]}
Critically examine the evolution of the Muslim League’s demand for Pakistan from 1940 to 1947.

\subsection*{Q.14 [Gandhi’s Role in the 1940s]}
Evaluate Gandhi’s role during the crucial decade leading up to Indian independence and Partition.

\subsection*{Q.15 [Transfer of Power]}
Was the transfer of power in 1947 a result of negotiations or a hurried British withdrawal? Discuss with arguments.
