\begin{center}
    \Large\textbf{UPSC Mains – GS Paper 1}\\
    \normalsize Chapter 3: \textit{Early Indian Responses — Reform and Rebellion}\\
    \vspace{0.5em}
    \textit{Based on \textbf{From Plassey to Partition} by Sekhar Bandyopadhyay}
\end{center}

\vspace{1em}

\subsection*{Q.1 [Social Reform Movements]}
Discuss the role of Raja Rammohan Roy in the emergence of social reform in 19th-century India.

\subsection*{Q.2 [Religious Reform and Modernity]}
To what extent did religious reform movements of the 19th century reflect the tension between tradition and modernity?

\subsection*{Q.3 [Brahmo Samaj and Arya Samaj]}
Compare and contrast the objectives and approaches of Brahmo Samaj and Arya Samaj.

\subsection*{Q.4 [Peasant Resistance]}
Examine the causes and consequences of major peasant revolts in early colonial India, such as the Indigo revolt and Pabna uprisings.

\subsection*{Q.5 [Tribal Uprisings]}
What were the major tribal uprisings during the early colonial period? Analyze their causes and leadership patterns.

\subsection*{Q.6 [Role of Religion in Rebellion]}
How did religion influence the mobilization during the Revolt of 1857?

\subsection*{Q.7 [Revolt of 1857: Nature]}
Was the Revolt of 1857 a sepoy mutiny, a civil rebellion, or the first war of independence? Justify your answer.

\subsection*{Q.8 [British Response to 1857]}
How did the British respond administratively and militarily after the Revolt of 1857?

\subsection*{Q.9 [Socio-Economic Background of 1857]}
Analyze the social and economic grievances that contributed to the Revolt of 1857.

\subsection*{Q.10 [Leadership in the Revolt]}
Evaluate the role of both military and civilian leadership in the Revolt of 1857.

\subsection*{Q.11 [Limits of the Revolt]}
What were the key reasons for the failure of the Revolt of 1857?

\subsection*{Q.12 [Intellectual Awakening]}
How did early Indian thinkers interpret colonial rule? Comment with reference to Derozio and the Young Bengal movement.

\subsection*{Q.13 [Women in Reform Movements]}
Discuss the role of women in 19th-century reform movements.

\subsection*{Q.14 [Education and Reform]}
How did Western education influence the reformist agenda in colonial India?

\subsection*{Q.15 [1857 in Historical Memory]}
How has the Revolt of 1857 been remembered in nationalist historiography?